---
title: "Homework Assignment 1"
author: Umang Arora
date: 'Assigned: Oct 24, 2020, Due Sun Nov 01, 2020 11:59PM'
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
---

## HW-1 Instructions

> Our class emphasizes clear communication of data analysis results to non-technical audiences. I expect your HTML output documents to be readable and well formatted. I expect you to put ample comments in your R code to make the code understandable. Along with accuracy of results, I am looking for quality of presentation as well. This homework is due by **11:59PM on Nov 01st**.  To complete this assignment, follow these steps:

1. Create a new RStudio Project for this HW. Sync the project directory with a GitHub Repository (see instructions on Canvas on how to do that). 

2. Download the `HW1.Rmd` file from Canvas. Save the file in the RStudio Project Directory. 

3. Open `HW1.Rmd` in RStudio. Replace the "Your Name Here" text in the `author:` field with your name.

4. Supply your solutions to the homework by editing `HW1.Rmd`.

5. Run your code in the Console and Knit HTML frequently to check for errors.

6. You may find it easier to solve a problem by interacting only with the Console at first. 

7. When you have completed the homework please check that your knits correctly when you click `Knit HTML`.

8. "Push" all your local changes to the GitHub Repo for the project.

9. Submit your RMD file, the HTML output and a link to your GitHub Repo on Canvas.


## Data frame basics

We will continue working with the nycflights dataset we looked at in class. **Please be sure to keep the data file in the same directory as the RMD file - your project directory.**

First - make sure that you are reading the data correctly and doing all the data cleaning steps that we did in class. Feel free to go beyond our class work in cleaning the data if you feel the need. 

```{r}
nyc <- read.csv("nycflights.csv")
#checking character types
str(nyc)
#converting carrier to factors 
nyc$carrier1<-as.factor(nyc$carrier)
```

I now would like you to answer the following - all in their own separate R code blocks.

### Data Exploration

Let's first do some simple exploration of this data. 

- How many airlines are there? (Hint: `levels` and `length` can be useful here)
```{r}
# cacluating # of unique carriers
airlines<- length(levels(nyc$carrier1))
airlines
```
- How many flights there were by the airline with code `OO`? (Hint: `nrow` can be useful here along with logical indexing)
```{r}
#code 00
flight_00<- nrow(nyc[nyc$carrier1 == 'OO', ])
flight_00
```
- How long is the shortest flight out of any NYC airport? (Hint: `min` can be useful, remember to handle `NA` values)
```{r}
#shortest flight
shortest_time <- min(nyc$air_time, na.rm=TRUE)
shortest_time
```
- How many flights where there by United Airlines (code: UA) on Jan 12th 2013?
```{r}
#flights by UA on Jan12th 2013
no_of_flights<- nrow(nyc[nyc$carrier == "UA" & nyc$month == 1 & nyc$year == 2013 & 
                           nyc$day==12, ])
no_of_flights
```

### Arrival Delay

Lets focus on Arrival Delay.

- What was the average arrival delay for all airports and all airlines combined in Jan 2013? 
```{r}
avg_delay <- mean(nyc[nyc$month == 1, ]$arr_delay, na.rm = TRUE)
avg_delay
```
- Whats was the median arrival delay for all airports and all airlines combined in Jan 2013?
```{r}
median_delay <- median(nyc[nyc$month == 1, ]$arr_delay, na.rm = TRUE)
median_delay
```

Based on your answers to the two questions above, what can you say about the distribution of arrival delays? Provide your answer in a text paragraph form.

- This implies that most of the flights are not delayed and actually reach before time. The flights that are late are delayed by a lot

### Airline Performance

Lets see if all airlines are equally terrible as far as flight arrival delays are concerned. For this question you will have to make sure that airline column is coded as a factor.

- Calculate average arrival delays by airline (Hint: look up the command `tapply`)
```{r}
avg_delay_carrier <- round(tapply(nyc$arr_delay, nyc$carrier, mean, na.rm = TRUE),2)
avg_delay_carrier
```

- Draw a Bar Plot of Average Arrival Delays for all the Airlines (Hint: command for making a Bar Plot is simply `barplot`)
```{r}
barplot(avg_delay_carrier)
```
- Which airline has the highest average arrival delay? Which airline has the smallest average arrival delay? Are there airlines that actually have negative average delay? Provide answer to this question in a text paragraph form using **inline R code**.
-- F9 has the highest average delay of `r max(avg_delay_carrier)`, AS has least delay of  `r min(avg_delay_carrier)`, HA also has negative delay along with AS

### Air Gain

Create a new column named airgain such that airgain = (departure delay - arrival delay) : this is the amount of delay a flight made up while in air.
```{r}
nyc$airgain <- nyc$dep_delay - nyc$arr_delay
```

a) Explore airgain data - calculate suitable descriptive statistics and appropriate graphics to better understand this data. This part is open ended - you do what you feel works best for you.
```{r}
hist(nyc$airgain)
```

b) Answer the questions:

  - do airlines actually gain any time when in air on average? 
  --The airgain is actually positive for most of the flights. So the flights actually gain time while flying

  - Calculate average airgain for different airlines - which airlines do a better job, which do a worse job? 
  ```{r}
  avg_airgain <- round(tapply(nyc$airgain, nyc$carrier, mean, na.rm = TRUE),2)
```
-- AS does the best job while F9 does the worst

### Merging Data Frames

> This section and the next is new compared to the class exercise. As you had an opportunity to work together in your breakout rooms for previous questions, this and the next section will carry a higher weight in grading for this HW.

You can get detailed information about the physical planes in our dataset in this file: `planes.csv`. Download and save this file in your project directory.

a) Read the `planes.csv` file using `read.csv` command. Do any data cleaning necessary.

```{r}
planes <- read.csv("planes.csv")
str(planes)
```

b) Merge the flights data and the planes data using the `merge` command. You should do the merge on the common column named `tailnum`. *getting this right may need some trial and error and getting some help*.

```{r}
new <- merge(nyc,planes, by='tailnum')
```


c) Now that you have a merged dataset, think of what interesting questions that you can ask that can be answered using the merged dataset. You are asked to pose five interesting questions and answer them. (For example: who are the top 10 manufacturers of planes that fly out of NYC airports?) **Be creative. Be bold. Ask questions that you would want to know answers to even if you were not doing this for a HW. **

Q1 Does engine type is related to delays
```{r}
avg_delay_engine <- round(tapply(new$arr_delay, new$engine, mean, na.rm = TRUE),2)
avg_delay_engine
```
-- It can be seen that that while Turbo jet and Turbo-prop has least delays, 4 cycle and turbo shaft has maximum delays

Q2. Top 10 manufacturers of planes that fly from NYC airports

```{r} 
#number of flights per manufacturer
list<- sort(table(new$manufacturer), decreasing = TRUE)
#taking top 10 manufacturers
head(list)
```
Q3 What is the type that is most flown out of nyc

```{r}
table (new$type)
```
-- MOst of them are multi engine fixed wing followed by single engine and rotorcraft. This makes sense since most are big commercial aircrafts followed by private jets maybe and then training or private small aircrafts. Let us see if single engine are more on time than multi engine

Q4 Are single engine planes more on time than large multi engine planes

```{r}
avg_delay_type <- round(tapply(new$arr_delay, new$type, mean, na.rm = TRUE),2)
avg_delay_type
```
-- While there is not a huge difference it seems that single engine planes has less delays of `r min(avg_delay_type)` while rotorcraft has maximum delay of `r max(avg_delay_type)`

Q5 Which carrier has larger capacity aircrafts on average
```{r}
avg_capacity_carrier <- round(tapply(new$seats, new$carrier, mean, na.rm = TRUE),2)
avg_capacity_carrier
```
-- HA has huge average capacity of 377 seats which is considerably higher than other airlines. MQ has smallest planes with average seating of 13 seats

### Making Your HTML Look Nice

We want our report to be good looking, professional documents. To that end, I am asking you to do the following:

- Have a floating table of contents

- Include code folding in your output. You can find more about code folding here: https://bookdown.org/yihui/rmarkdown/html-document.html#code-folding


That's it. Once you are done, make sure everything works and knits well and then you can push your changes to the GitHub repo and uplaod the RMD flile and the html output to Canvas.

**Have Fun!**

Sanjeev
